---
title: "HW2 STA521 Fall18"
author: 'Andrew Cooper, ahc15, andrewdjac'
date: "Due September 23, 2018 5pm"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

## Backgound Reading

Readings: Chapters 3-4 in Weisberg Applied Linear Regression


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This exercise involves the UN data set from `alr3` package. Install `alr3` and the `car` packages and load the data to answer the following questions adding your code in the code chunks.  Please add appropriate code to the chunks to suppress messages and warnings as needed once you are sure the code is working properly and remove instructions if no longer needed. Figures should have informative captions. Please switch the output to pdf for your final version to upload to Sakai. **Remove these instructions for final submission**


## Exploratory Data Analysis

```{r data, echo=F, message=F, warning=F}
library(alr3)
data(UN3, package="alr3")
help(UN3) 
library(car)
library(knitr)
library(ggplot2)
library(GGally)
library(gridExtra)
```


### Problem 1

```{r}
summary(UN3)
```

As we can see in the summary, six of the variables in the UN3 dataset have missing data. In addition, all of the variables in the dataset are quantitative.

### Problem 2

To get a better sense of the variables in the UN3 dataset, we create a table showing each variable's mean and standard deviation.

```{r, message=F, warning=F}
sum_tab <- apply(X=UN3,MARGIN=2,FUN=function(x){c(mean(x,na.rm=T),sd(x,na.rm=T))})
sum_tab <- t(sum_tab)
colnames(sum_tab) <- c("mean", "standard deviation")
kable(sum_tab, caption="Means and standard deviations of variables in UN3")
```


### Problem 3

```{r,message=F,warning=F}
ggpairs(UN3,title="Pairwise relationships of variables in UN3")
```

The ggpairs plot above gives us a lot of useful information regarding the predictors and their relationships. First, the diagonal plots showing each predictor's distribution reveals the variables "PPgdp", "Change", and "Fertility" are all heavily right-skewed. This is one indication that perhaps a log transformation might be appropriate for these variables. In particular, the plots for row "Pop" are all hard to decipher since most of the values are close to zero. This could perhaps be accounted for by logging "Pop" so that it's more normally-distributed. 

Second, the pairs plot shows linear-looking relationships between "ModernC" and "Change", "ModernC" and "Frate", "PPgdp" and "Frate", "ModernC" and "Fertility", "Change" and "Fertility", "ModernC" and "Purban", "Frate" and "Purban", and "Fertility" and "Purban". In addition, see potentially non-linear relationships between "ModernC" and "PPgdp", "PPgdp" and "Fertility", and "PPgdp" and "Purban". These relationships look like log or perhaps square-root relationships.   

To get a better look at the relationships between the predictors and our response "ModernC", we create individual scatterplots below.

```{r,message=F,warning=F}
p1 <- qplot(data=UN3,x=Change,y=ModernC)+labs(title="Change vs. ModernC")
p2 <- qplot(data=UN3,x=PPgdp,y=ModernC)+labs(title="PPgdp vs. ModernC")
p3 <- qplot(data=UN3,x=Frate,y=ModernC)+labs(title="Frate vs. ModernC")
p4 <- qplot(data=UN3,x=Pop,y=ModernC)+labs(title="Pop vs. ModernC")
p5 <- qplot(data=UN3,x=Fertility,y=ModernC)+labs(title="Fertility vs. ModernC")
p6 <- qplot(data=UN3,x=Purban,y=ModernC)+labs(title="Purban vs. ModernC")
grid.arrange(p1, p2, p3, p4, p5, p6, nrow = 3)
```

"Change" appears to have a negative relationship with "ModernC", although it is difficult to tell if this relationship linear or perhaps non-linear due to the slight curve in the scatterplot.

It is difficult to evaluate the relationship between "PPgdp" and "ModernC" since most of the values in "PPgdp" are clustered around zero. As stated earlier, "PPgdp" is heavily right-skewed, indicating a log-transformation might be appropriate. 

"Frate" does not appear to have any apparent relationship with "ModernC". Perhaps there's a slightly positive relationship, but it is weak as best.


It is difficult to evaluate the relationship between "Pop" and "ModernC" since most of the values in "Pop" are clustered around zero. As stated earlier, "Pop" is heavily right-skewed, indicating a log-transformation might be appropriate. In addition, there are two outliers on the right that are likely values of large leverage. However, those points might not be as much of outliers if a log transformation is performed on "Pop".

"Fertility" has a rather strong negative relationship with "ModernC". It is difficult to tell if the relationship is strictly linear or has some slight curvature to it.

Finally, "Purban" has a positive and linear-looking relationship with "ModernC".

## Model Fitting

### Problem 4

```{r}
par(mfrow=c(2,2))
lm1 <- lm(ModernC~Change+PPgdp+Frate+Pop+Fertility+Purban,data=UN3)
plot(lm1)
```

The residuals appear to be approximately evenly distributed above and the below the line and showing no obvious pattern, indicating the linear fit is appropriate and the residuals are independent. There might be more variance in the center of the plot than the left and right sides, which could violate the equal variance condition, but for the most part the residuals look good. The qq-plot shows pretty normal-looking standardized residuals, satisfying the normality condition. There are some outliers of note. Azerbaijan, Poland, and the Cook Islands have particularly large residuals, indicating a lack of fit. In addition, Kuwait, India, and China have particularly large leverage. 

```{r}
length(lm1$residuals)
```

Printing the number of residuals in our model, we see the model fitting used 125 of our observations. Those are the number of "complete" cases in our UN3 dataset, i.e. all the observations with no missing entries. This means the lm() function automatically threw out all the incomplete cases before fitting the model.

### Problem 5 

```{r,message=F,warning=F}
car::avPlots(lm1,data=UN3)
```

We observe outliers and potentially influential points in both the added variable plots for "PPgdp" and "Pop". In the "PPgdp" plot, the outliers are the countries Switzerland and Norway. In the "Pop" plot the outliers are China and India. These two plots suggest that perhaps a log transformation is needed for those terms. A log transformation might make these points less influential.

The other plots look rather linear with no influential points.

### Problem 6

We first decide to make "PPgdp" and "Pop" candidates for transformation in our boxTidwell() function, as reasoned by our analysis of the added variable plots.

```{r}
boxTidwell(ModernC~PPgdp+Pop, ~ Change+Frate+Fertility+Purban, data=UN3)
```

The non-signficant lambda values indicate the power transformations are not needed for these predictors.

We found in previous analysis reasons to perhaps use the log-transformation on the predictors "PPgdp", "Pop", and "Fertility". So we now log-transform these predictors and see how their relationships with "ModernC" change.

```{r, message=F, warning=F}
p1 <- qplot(data=UN3,x=log(PPgdp),y=ModernC)+labs(title="log(PPgdp) vs. ModernC")
p2 <- qplot(data=UN3,x=log(Pop),y=ModernC)+labs(title="log(Pop) vs. ModernC")
p3 <- qplot(data=UN3,x=log(Fertility),y=ModernC)+labs(title="log(Fertility) vs. ModernC")
grid.arrange(p1, p2, p3, nrow = 2)
```

The scatter plots indicate that both predictors, particularly "PPgdp", benefit from a log transformation, as their relationships with "ModernC" now look more linear.

We decide to only log-transform the predictors "PPgdp", "Pop", and "Fertility" in our model.

### Problem 7


```{r, message=F, warning=F}
boxCox(lm(ModernC~Change+log(PPgdp)+Frate+log(Pop)+log(Fertility)+Purban, data=UN3))
```
 
The confidence interval for Lambda produced by the boxCox() function include the value 1, indicating a transformation of the response might not be signficantly better than no transformation. We therefore decide a transformation of the response "ModernC" is not necessary. 
 
### Problem 8

```{r, message=F, warning=F}
par(mfrow=c(2,2))
lm2 <- lm(ModernC~Change+log(PPgdp)+Frate+log(Pop)+log(Fertility)+Purban,data=UN3)
summary(lm2)
```

The coefficients for "Change" and "Purban" are not signficant according to the summary, indicating they don't provide enough information in predicting "ModernC". We remove these predictors and refit the model.

```{r, message=F, warning-F}
lm3 <- lm(ModernC~log(PPgdp)+Frate+log(Pop)+log(Fertility),data=UN3)
summary(lm3)
```

Now all the predictors are significant and the R-squared value is about the same, so we choose this as the final model.

```{r, message=F, warning=F}
par(mfrow=c(2,2))
plot(lm3)
```


The residuals appear randomly and evenly distributed above and below the fitted line, indicating a linear relationship is appropriate and that the residuals are indendent. The qq-plot also has small tails indicating normally distributed standardized residuals. 

```{r, message=F, warning=F}
car::avPlots(lm3,data=UN3)
```

All the added variable plots look linear with no significant outliers or influential points.

### Problem 9

We first use the boxcox() function to find the optimal response transformation.

```{r, message=F, warning=F}
boxCox(lm(ModernC~Change+PPgdp+Frate+Pop+Fertility+Purban, data=UN3))
```

Based on the confidence interval for lambda, which includes the value 1, we decide to not use a transformation on the response.

We then run the boxTidwell() function to decide which predictors to transform.

```{r, message=F, warning=F, eval=F}
boxTidwell(ModernC~PPgdp+Pop+Change+100+Frate+Fertility+Purban, data=UN3)
```

The code in the above chunk failed to diverge and returned an error. This seems to be due to a bug in the boxTidwell package function. For this reason, we decide the appropriate transformations of the predictors from graphical analysis.

As we showed in previous analysis, the predictors "PPgdp", "Pop", and "Fertility" are heavily right-skewed, non-negative variables. Their relationships with the response "ModernC" is hard to evaluate given how all the values are clustered around zero, so a log transformation seems appropriate. Taking the log of these predictors yields more normal-looking distributions and more linear-looking relationships with "ModernC".

After log-transforming "PPgdp", "Pop", and "Fertility", we get the final model

$$
E[ModernC|X] = \beta_0 + \beta_1Change + \beta_2log(PPgdp) + \beta_3Frate + \beta_4log(Pop) + \beta_5log(Fertility) + \beta_6Purban
$$
This happens to be the same model we constructed in problem 8.


### Problem 10

Looking at the residual diagnostic plots for the final model, it appears the Cook Islands, China, Armenia, Nicaragua, and Poland are outliers in the data. We remove these from the data and refit the model.

```{r, message=F, warning=F}
UN3_new <- UN3[!(rownames(UN3) %in% c("Cook.Islands","China","Armenia", "Nicaragua", "Poland")),]
lm4 <- lm(ModernC~Change+log(PPgdp)+Frate+log(Pop)+log(Fertility)+Purban,data=UN3_new)
par(mfrow=c(2,2))
plot(lm4)
```

Looking at the residual plots, there are new outliers but they have little leverage. However, the points we removed were not egregious outliers such that they were unfit for our model. None of the removed points were within Cook's distance, so we decide to put the points back in.

## Summary of Results

### Problem 11


```{r}
int_tab <- as.data.frame(confint(lm3))
int_tab$Coefficient <- lm3$coefficients
rownames(int_tab) <- c("Intercept","PPgdp","Frate","Pop","Fertility")
kable(int_tab, caption="Coefficient estimates and confidence intervals for final model")
```

Intercept: When all predictors are zero, the expected percent of unmarried women using modern contraception is -19.47%. This coefficient is clearly doesn't make sense in the context of the problem.

PPgdp: When the 2001 GDP increases by a factor of 10, the expected percentage of unmarried women using modern contraception increases by 6.63%.

Frate: For every percent increase in females over 15 who are economically active, the expected percentage of unmarried women using modern contraception increases by 0.16%.

Pop: When the population increase by a factor of 10, the expected percentage of unmarried women using modern contraception increases by 1.77%

Fertility: When the fertility rate increases by a factor of 10, the expected percentage of unmarried women using modern contraception decreases by 13.62%.


### Problem 12


Our final model is

$$
\hat{ModernC} = -19.47 + 6.63og(PPgdp) + 0.16(Frate) + 1.77log(Pop) - 13.62log(Fertility) 
$$

The model gives us insight into the relationships between general features of the female UN population and the use of modern methods of contraception. The coefficient for "PPgdp" implies as a country's GDP increases, so does its percentage of women using modern contraception. This could be because countries with higher GDP's tend to also be more developed both socially and technologically. In these countries, modern contraception might be more available or standard than less developed countries with lower GDP's.

The coefficient for "Frate" indicates that as more women become economically active, more women use modern contraception. This makes sense as well, as countries with more economically active women typically have more rights for women and are more socially progressive, which would lead to more women using modern contraception. On the other hand, countries with less economically active women tend to have fewer women's rights, so using modern contraception might be stigmatized or even illegal.  

The coefficient for "Pop" implies that as the population increases, so does the percentage of women using modern contraception. This could be because countries with higher populations are typically more developed and technologically advanced, and therefore are more likely introduced to modern methods of contraception. 

Finally, the coefficient for Fertility implies that as fertility rates increase, the percentage of women using modern contraception decreases. This makes sense, as fertility rate and use of contraception are directly related. Countries with higher fertility rates have more expected children per female, indicating women in that country tend to use fewer methods of contraceptives. Likewise, countries with lower fertility rates likely have women who often use contraceptives to avoid pregnancy.

## Methodology

    
### Problem 13

The formula for the added variable regression line is 

$$
\hat{e_y} = \hat{\beta_0}+\hat{\beta_1}\hat{e_{x_i}}
$$
Where $\hat{e_y}$ are the residuals of the model regressing $Y$ against all the predictors except for $x_i$, and $\hat{e_{x_i}}$ are the residuals of the model regressing $x_i$ against all the predictors except for $x_i$. We can rewrite this as

$$
(I-H)Y = \hat{\beta_0}1_{nx1}+\hat{\beta_1}(I-H)x_3
$$
Where $1_{nx1}$ is a vector of 1's. We then substitute for $\hat{\beta_1}$ the formula for the OLS estimate:

$$(I-H)Y = \hat{\beta_0}1_{nx1} + [((I-H)x_i)^T((I-H)x_i)]^{-1}((I-H)x_i)^T(I-H)Y(I-H)x_i$$
$$(I-H)Y = \hat{\beta_0}1_{nx1} + [x_i^T(I-H)^T(I-H)x_i]^{-1}x_i^T(I-H)^T(I-H)Y(I-H)x_i $$
$$(I-H)Y = \hat{\beta_0}1_{nx1} + [x_i^T(I-H)x_i]^{-1}x_i^T(I-H)Y(I-H)x_i$$

We then multiply each side by $x_i^T$ to simplify the expression

$$x_i^T(I-H)Y = x_i^T\hat{\beta_0}1_{nx1} + x_i^T[x_i^T(I-H)x_i]^{-1}x_i^T(I-H)Y(I-H)x_i $$
$$x_i^T(I-H)Y = \hat{\beta_0}\sum{x_i}+x_i^T(I-H)x_i[x_i^T(I-H)x_i]^{-1}x_i^T(I-H)Y $$
We see a term and its inverse multiplied together on the right-hand side, which can be simplified to get

$$x_i^T(I-H)Y = \hat{\beta_0}\sum{x_i}+x_i^T(I-H)Y $$
$$\hat{\beta_0}\sum{x_i} = 0 $$
$$\hat{\beta_0} = 0$$

### Problem 14

We show that the estimate for the coefficient for "Purban" is the same using both methods.

First, we get the coefficient estimate from the full model we created earlier.

```{r, message=F, warning=F}
lm2$coefficients[7]
```

Then we construct two new models. The first model takes our full model and takes out "Purban" as a predictor. The second model takes that model and replaces the response with "Purban". The residuals of the first model ($e_Y$) are regressed against the residuals of the second model ($e_X$):

```{r, message=F, warning=F}
UN3_complete <- UN3[complete.cases(UN3),]
lm4 <- lm(ModernC ~ Change + log(PPgdp) + Frate + log(Pop) + log(Fertility), data=UN3_complete)
lm5 <- lm(Purban ~ Change + log(PPgdp) + Frate + log(Pop) + log(Fertility), data=UN3_complete)
e_Y <- lm4$residuals
e_X <- lm5$residuals
lm6 <- lm(e_Y ~ e_X)
lm6$coefficients[2]
```

As we can see, the two estimates for the coefficient for "Purban" are the same ($\approx-.00735$).